---
title: "TCGA_mRNA_script"
author: "Tara Morrison"
date: "05/11/2018"
output: word_document
editor_options: 
  chunk_output_type: console
---
##Initial set up
Working Directory (Lab) & Libraries
``` {r echo=FALSE, r eval=F}
setwd("~/Desktop/part2_project/Liver_Cancer")

library(RColorBrewer)
library(gplots)
library(ggplot2)
library(DESeq2)
library(Rtsne)
library(pca3d)
library(limma)
library(genefilter)
library(sva)
library(edgeR)
library(statmod)
library(reticulate)
library(VennDiagram)

```

##TCGA DATA

#Upload pdata (TCGA)
```{r}
#pdata=read.table('~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/pdata_rnaseq_genelevel_small.txt', header=T)
#rownames(pdata)=pdata$Sample

#Scaling up to full TCGA data set
pdata=read.table('~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/pdata_rnaseq_genelevel.txt', header=T)
rownames(pdata)=pdata$Sample
pdata = pdata[pdata$Project_ID != "TCGA-CHOL",]
pdata=pdata[pdata$Sample_Type != 'Recurrent_Tumor',]
pdata$Project_ID = droplevels(pdata$Project_ID)
pdata$Sample_Type=droplevels(pdata$Sample_Type)
```

#Set Conditions/Sample Types
There are 3 levels: Primary Tumour, Recurrent Tumour and Normal Tissue
```{r}

#Identify the conditions and allocate a colour to each
conds=(pdata$Sample_Type)
cond_colours=brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
```

#Read in count data
Data taken from TCGA data repository - Liver Cancer mRNA
datatype=HTSeq count data
```{r}

##TCGA DATA COUNTS
#Count data is loaded in from HTSeq Count available in the pdata file retreived from TCGA

ddsHTSeq=DESeqDataSetFromHTSeqCount(sampleTable = pdata, directory= '~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel', design = ~ Sample_Type)

# These gene IDs have extra numbers at the end - folowing code used to remove
rownames(ddsHTSeq) = gsub(".\\d+$","",rownames(ddsHTSeq),perl=T)



colData(ddsHTSeq)$Sample_Type=factor(colData(ddsHTSeq)$Sample_Type, levels=levels(pdata$Sample_Type))

dds=estimateSizeFactors(ddsHTSeq)
dds=estimateDispersions(dds)

```

#Data Normalisation
```{r}

normcounts = counts(dds, normalized=TRUE)
tcgarawcounts=counts(dds,normalized=FALSE) #These are used later when merging with GTEx
colnames(tcgarawcounts)=pdata$Sample_Type
log2counts=log2(normcounts+1)

```

#Quality Control
#Normalisation is generally ok. Consistent differences between both tumour tissues and normal tissues post-normalisation.
```{r}

quartz()
par(mfrow=c(1,2))

barplot(colSums(counts(dds, normalized=FALSE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Pre Normalised Counts - TCGA', names.arg =conds)
legend("topleft",levels(conds),cex=0.5,fill=cond_colours)

barplot(colSums(counts(dds, normalized=TRUE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Post Normalised Counts - TCGA', names.arg =conds)
legend("topleft",levels(conds),cex=0.5,fill=cond_colours)

```

#Dispersion Plot
```{r echo = FALSE}

dds=nbinomWaldTest(dds)

counts_table=counts(dds, normalized=TRUE)

quartz()
plotDispEsts(dds,main='Dispersion Plot - TCGA')

```


##Variance Stabilising Transformation
```{r}
vsd <- varianceStabilizingTransformation(dds)
vstMat <- assay(vsd)
vstcounts <- vstMat[order(apply(vstMat,1,sum),decreasing =TRUE),]

quartz()

heatmap.2(cor(vstcounts),trace="none",labRow = conds, cexRow = 0.6, labCol = pdata$Sample_Type, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the VST Counts - TCGA', Rowv = F, ColSideColors = cond_colours[conds])
```


#PCA Plots & Heatmaps
#PCA plots shows good separation between tumour tissues and solid normals. One primary tumour point appears to be anomalous whilst another closer to recurrent tumour tissue type.
#Heatmap shows good clustering of solid tissue normals. Heatmap supports PCA plot in that one of the primary tumour points appears to be anomalous and does not closely match other primary tumours whilst another more closely matches recurrent tumour tissues.
```{r echo = FALSE}

#ggplot PCA
pca = prcomp(t(assay(vsd)))

pca.df = pca$x[, c(1,2,3)] #* -1 # if necessary flip coords for better visualisation
# Percentage explained by each component
percentages = round(pca$sdev^2 / sum(pca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(pca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")
pca.df = data.frame(pca.df, Condition = dds$Sample_Type)

quartz()
ggplot(pca.df, aes(x=PC1,y=PC2, colour = Condition))+
	geom_point(size = 4)+
  scale_color_manual(values=cond_colours[conds])+
  ggtitle("PCA1 vs PCA2 (TCGA - DESeq)")+
  stat_ellipse(aes(colour= pca.df$Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-200,220) +
  ylim(-200, 210)


#3D PCA Plot with 2D comparison
pca3d(pca, group=conds, col=cond_colours, legend = 'bottomright')

quartz()
pca2d(pca, group=conds, col=cond_colours, legend='bottomright')


```


##Rtsne

```{r}

#Normalised counts data into tsne object
set.seed(72)
tsne = Rtsne(t(vstMat), perplexity = 3)

#convert to data frame with 2 dimensions
tsne.df=data.frame(tsne.1=tsne$Y[,1], tsne.2=tsne$Y[,2], Type=as.factor(dds$Sample_Type))
quartz()
ggplot(data = tsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Type))+
  geom_point(size=4)+
  ggtitle("T-sne (TCGA - DESeq)")+
  scale_color_manual(values=unique(cond_colours[tsne.df$Type]))+
  stat_ellipse(aes(colour= tsne.df$Type),type = "t")+
  xlim(-55,60) +
  ylim(-60, 60)

```

#Statistical Analysis

```{r}

#Threshold for P-value is set to 0.05 and log fold change of 1.5 (absolute lfc used)
#For loop in place to consecutively compare primary and recurrent tumour against normals
#Signifciant hits identified using the adjusted p value and the log fold change - producing volcano plots and heatmap

p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250

condition_list=levels(conds)

for (i in 2:(length(condition_list))){
 
  normal=condition_list[1]
  compare=condition_list[i]

  print(paste('Comparing', normal, 'vs.', compare))
  
  res=results(dds, contrast = c('Sample_Type', normal, compare))
  res=res[order(abs(res$log2FoldChange), decreasing = T),]
  
  hits=(rownames(res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))

   print(paste("Found: ",length(hits)," Statistical hits"))
   
  if (length(hits) != 0){
    if (length(hits)>max_display_hits){
      hits=hits[1:max_display_hits,]
    }
    
  quartz()
  plot(res$log2FoldChange,-log10(res$padj), ylab='-log10(Adjusted P-value)', xlab='log2 FoldChange', main=paste('Volcano Plot:', normal, 'vs.', compare), pch=19, cex=0.2)
  points(res[hits,'log2FoldChange'],-log(res[hits,'padj'],10),pch=21,cex=0.4,col='turquoise1')
  text(res[hits[1:100],]$log2FoldChange,-log10(res[hits[1:100],]$padj), labels = gene_names[hits[1:100],]$GeneName, pos=3, cex = 0.4)
  abline(h=-log10(0.05), lty=3, col='red')
  abline(v=-1, lty=3, col='red')
  abline(v=1, lty=3, col='red')
 
  quartz()
  heatmap.2(vstMat[hits,],trace='none',col=hmcol,Colv=FALSE,dendrogram='row',main=paste('Heatmap Sig. Hits for', compare),labCol = conds, cexCol = 0.5, ColSideColors = cond_colours[conds]) 
  } else {
    print(paste("No Results for ",compare))
  }
  
}

```

```{r}

hits=((res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
upregulated=hits[hits$log2FoldChange>0,]
downregulated=hits[hits$log2FoldChange<0,]

```
##Median
```{r echo=FALSE}

#Median found across the rows for each condition
median.primary=apply(vstMat[,1:371],1,median)
median.normal=apply(vstMat[,372:421],1,median)

quartz()
plot(median.normal,median.primary, main="Normal vs. Primary_Tumor", pch=19, col='darkblue', cex=0.4)
points(median.normal[hits],median.primary[hits], pch=21, col='turquoise1')
text(median.normal[hits],median.primary[hits],cex=0.4,pos=4,labels = gene_names[hits,]$GeneName)
abline(a=0,b=1, col='red', lwd=2, cex=0.4, lty=2)


```

```{r}
#Running loop sepcifying [2] for Primary Tumour and [3] for Recurrent Tumour to save out hits
TCGAprimhits=as.data.frame(gene_names[hits[1:250],]$GeneName)
```


#Biomarkers things?
```{r}

biomarkers=vstcounts[c('ENSG00000081051','ENSG00000147257', 'ENSG00000110492', 'ENSG00000135052', 'ENSG00000118785'),]

quartz()

heatmap.2(biomarkers, trace='none', Colv = F, ColSideColors = cond_colours[conds], labRow = gene_names[c(rownames(biomarkers)),]$GeneName, main='Heatmap of VST counts for HCC Biomarkers')

```

#Save out plots & workspace
```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/TCGA_only.RData")
```

##Voom

```{r}

dge=DGEList(counts=tcgarawcounts)

dge$samples$group=conds

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()
par(mfrow=c(2,1))
barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

lcpm=cpm(dge, log=T)

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data - Voom",ylab="Log-cpm", cex.axis = 0.5)

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - TCGA (Voom)', Rowv = F, ColSideColors = cond_colours[conds])
```

#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)

#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

#PCA & Tsne
```{r}
#PCA Plot
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)] #* -1 # if necessary flip coords for better visualisation
# Percentage explained by each component
percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")
voompca.df = data.frame(voompca.df, Condition = dge$samples$group)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (TCGA -Voom)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-200,220) +
  ylim(-200, 210)


#Tsne
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type))+
  geom_point(size=4)+
  stat_ellipse(aes(colour=Type),type = "t")+
  ggtitle("T-sne (TCGA-Voom)")+
  scale_color_manual(values=unique(cond_colours[conds])) +
  xlim(-30,30) +
  ylim(-15,15)
```


##Significant Hits
```{r}
#Extract top significant hits by adjusted p value and log fold change
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])


quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Voom)')
  points(hit_table1[hitlist1,"logFC"],-log10(hit_table1[hitlist1,"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', Colv=FALSE, col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = cond_colours[conds])

```

```{r}

VOOMprimhits=as.data.frame(gene_names[hitlist1[1:250],]$GeneName)

```

```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/TCGAvoom.Rdata")
```

#SVA Design

#Removing hidden batch effects using SVA
```{r}
DESeqdds=DESeq(dds)
dat=counts(DESeqdds,normalized=T)
idx=rowMeans(dat)>1
dat=dat[idx,]

mod=model.matrix(~Sample_Type, colData(DESeqdds))
mod0=model.matrix(~1, colData(DESeqdds))

n.sv = num.sv(dat,mod,method="leek") 

svseq=svaseq(dat, mod, mod0, n.sv = (n.sv-2))

quartz()
par(mfrow=c(3,1))

stripchart(svseq$sv[,1] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV1")
abline(h=0)

stripchart(svseq$sv[,2] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV2")
abline(h=0)

stripchart(svseq$sv[,3] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV3")
abline(h=0)


ddssva = DESeqdds

ddssva$SV1= svseq$sv[,1] 

design(ddssva)= ~ SV1 + Sample_Type

```


#Data normalisation - SVA
```{r}
svanormcounts <- counts(ddssva, normalized=TRUE)
svarawcounts=counts(ddssva,normalized=FALSE)
svalog2counts=log2(svanormcounts+1)

quartz()
par(mfrow=c(2,1))

barplot(colSums(counts(ddssva, normalized=FALSE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Pre Normalised Counts')
legend("topright",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])

barplot(colSums(counts(ddssva, normalized=TRUE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Post Normalised Counts - SVA')
legend("topleft",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])

```

```{r}
ddssva=nbinomWaldTest(ddssva)

counts_table=counts(dds, normalized=TRUE)

quartz()
plotDispEsts(dds,main='Dispersion Plot - GTEx')
```

#VST

#VST Normalisation of SVA
```{r}
svavsd <- varianceStabilizingTransformation(ddssva, blind = F)
svavstMat <- assay(svavsd)
svavstcounts <- svavstMat[order(apply(svavstMat,1,sum),decreasing =TRUE),]

quartz()
heatmap.2(cor(svavstcounts),trace="none",labRow = conds, cexRow = 0.6, labCol = conds, cexCol = 0.6, col=hmcol, main = 'Heatmap of VST counts- SVA', ColSideColors = cond_colours[conds])
```

#PCA Plots & Heatmaps - SVA
#Prcomp PCA
```{r}
svapca = prcomp(t(assay(svavsd)))

svapca.df = svapca$x[, c(1,2,3)] #* -1 # if necessary flip coords for better visualisation
# Percentage explained by each component
percentages = round(svapca$sdev^2 / sum(pca$sdev^2) * 100)[1:3]
percentages = paste(paste(colnames(svapca.df), ":"),  paste(as.character(percentages), "%"), "variance", sep=" ")
svapca.df = data.frame(svapca.df, Condition = ddssva$Sample_Type)

quartz()
ggplot(svapca.df, aes(x=PC1,y=PC2, colour = Condition)) + 
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (TCGA - SVA)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-200, 220) +
  ylim(-200, 210)

```

#Tsne
```{r}
svatsne = Rtsne(t(svavstMat), perplexity = 30)
svatsne.df=data.frame(tsne.1=svatsne$Y[,1], tsne.2=svatsne$Y[,2], Type=as.factor(ddssva$Sample_Type))

quartz()
ggplot(data = svatsne.df, aes(tsne.1,  tsne.2, colour=Type))+
  geom_point(size=4)+
  stat_ellipse(aes(colour=Type),type = "t")+
  scale_color_manual(values=unique(cond_colours[svatsne.df$Type]))+
  ggtitle("T-sne (TCGA - SVA)")+
  xlim(-30, 30) +
  ylim(-20, 20)
```

#Statistical Analysis - SVA
```{r}
condition_list=levels(conds)

for (i in 2:(length(condition_list))){
 
  normal=condition_list[1]
  compare=condition_list[2]

  print(paste('(SVA) Comparing', normal, 'vs.', compare))
  
  res=results(ddssva, contrast = c('Sample_Type', normal, compare))
  res=res[order(abs(res$log2FoldChange), decreasing = T),]
  
  hits=(rownames(res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
  
   print(paste("Found: ",length(hits)," Statistical hits"))
   
  if (length(hits) != 0){
    if (length(hits)>max_display_hits){
      hits=hits[1:max_display_hits]
    }
    
  quartz()
  plot(res$log2FoldChange,-log10(res$padj), ylab='-log10(Adjusted P-value)', xlab='log2 FoldChange', main=paste('Volcano Plot:', normal, 'vs.', compare,'(SVA)'), pch=19, cex=0.2)
  points(res[hits,'log2FoldChange'],-log(res[hits,'padj'],10),pch=21,cex=0.4,col='turquoise1')
  text(res[hits[1:100],]$log2FoldChange,-log10(res[hits[1:100],]$padj), labels = gene_names[hits[1:100],]$GeneName, pos=3, cex = 0.4)
  abline(h=-log10(0.05), lty=3, col='red')
  abline(v=-1, lty=3, col='red')
  abline(v=1, lty=3, col='red')

  
quartz()
  heatmap.2(svavstMat[hits,],trace='none',col=hmcol, dendrogram='col',main=paste("Heatmap Sig. Hits for", compare),labCol = conds, cexCol = 0.5, ColSideColors = cond_colours[conds]) 
  
  } else {
    print(paste("No Results for ",compare))
  }
  
}

quartz()

heatmap.2(log2(counts(ddssva,normalized=T)[rownames(res[1:250,]),]+1),col=hmcol,trace="none", ColSideColors = cond_colours[conds])
```

```{r}

SVAprimhits=as.data.frame(gene_names[hits,]$GeneName)

```

```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/TCGAsva.Rdata")
```
