---
title: "additional_analysis"
author: "Tara Morrison"
date: "07/03/2019"
output: word_document
---

```{r}
library(DropletUtils)
library(RColorBrewer)
library(gplots)
library(ggplot2)
library(DESeq2)
library(Rtsne)
library(pca3d)
library(limma)
library(genefilter)
library(sva)
library(edgeR)
library(statmod)
library(reticulate)
library(VennDiagram)
library(gridExtra)
library(gridGraphics)

```


##Venn Diagrams
#Using significant up and down regualted genes from Voom normalisation methods as selected when downsampling.
```{r}

tcga_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/tcgavoom_up.RDS')
tcga_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/tcgavoom_down.RDS')

merge_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/mergedvoom_up.RDS')
merge_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/mergedvoom_down.RDS')

gtex_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/gtexvoom_up.RDS')
gtex_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/gtexvoom_down.RDS')

colnames(tcga_up)='GeneName'
colnames(tcga_down)='GeneName'
colnames(merge_up)='GeneName'
colnames(merge_down)='GeneName'
colnames(gtex_up)='GeneName'
colnames(gtex_down)='GeneName'

voom.venn.up =  draw.triple.venn(

	area1 = nrow(tcga_up), 
	area2 = nrow(merge_up), 
	area3 = nrow(gtex_up),
	n12 = length(intersect(tcga_up$GeneName, merge_up$GeneName)),
	n23 = length(intersect(tcga_up$GeneName, gtex_up$GeneName)),
	n13 = length(intersect(merge_up$GeneName, gtex_up$GeneName)),
	n123 = length(intersect(intersect(tcga_up$GeneName, merge_up$GeneName), gtex_up$GeneName)),
	fill = c("blue", "red", "green"),
	category = c('TCGA', 'Merged', 'GTEx')

)

pdf(file='~/Desktop/part2_project/Liver_Cancer/plots/upregulatd_venn.pdf')
  grid.draw(voom.venn.up)
dev.off()


voom.venn.down =  draw.triple.venn(

	area1 = nrow(tcga_down), 
	area2 = nrow(merge_down), 
	area3 = nrow(gtex_down),
	n12 = length(intersect(tcga_down$GeneName, merge_down$GeneName)),
	n23 = length(intersect(tcga_down$GeneName, gtex_down$GeneName)),
	n13 = length(intersect(merge_down$GeneName, gtex_down$GeneName)),
	n123 = length(intersect(intersect(tcga_down$GeneName, merge_down$GeneName), gtex_down$GeneName)),
	fill = c("blue", "red", "green"),
	category = c('TCGA', 'Merged', 'GTEx')

)

pdf(file='~/Desktop/part2_project/Liver_Cancer/plots/downregulatd_venn.pdf')
  grid.draw(voom.venn.down)
dev.off()

save.image(file = '~/Desktop/part2_project/Liver_Cancer/updown.Rdata')
```



## Python downsampling code
#USing Voom method and downsampled counts.

#25% Downsample

#Load in original gtex matrix and downsample by 25%
```{r}
gtex=readRDS('GTEX.rds')

gtex_samples=read.table('gtex_simple_samples.txt',row.names=1,sep="\t",header=F)
colnames(gtex_samples)=c('Tissue_Type', 'Description')

#Liver samples only - '-' substituted for '.' to match initial GTEX counts data
liver_ids = rownames(gtex_samples)[gtex_samples$Tissue_Type == "Liver"]
liver_ids = gsub("-",".",liver_ids)

gtex = gtex[,intersect(liver_ids,colnames(gtex))]

rownames(gtex)=gsub(".\\d+$","",rownames(gtex),perl=T)

gtex=as.matrix(gtex)
gtex25=downsampleMatrix(gtex, 0.75)
```

#Load in tcgarawcounts and merge the two
```{r}
tcgarawcounts=readRDS('~/Desktop/part2_project/Liver_Cancer/tcgarawcounts.RDS')

common_ids = intersect(rownames(gtex25),rownames(tcgarawcounts))

gtex25  = gtex25[common_ids, ]

rawcounts = tcgarawcounts[common_ids, ] 

p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250
```

#MERGE GTEX Normals & TCGA Primary Tumour Matrices
```{r}
mergedcounts=cbind(gtex25, rawcounts, deparse.level=1)
mergedcounts=mergedcounts[rowSums(mergedcounts)>=3,]

#Labelling columns by Sample Type
colnames(mergedcounts)[1:ncol(gtex25)]='Normal'

#Identify Conditions and Batches (GTEX vs. TCGA)
conds=as.factor(colnames(mergedcounts))
cond_colours = brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

batch = as.factor(c(rep("GTEX",ncol(gtex)),rep("TCGA",ncol(rawcounts))))
batch_colours=c('greenyellow', 'dark green')
names(batch_colours)=unique(batch)

hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
```

```{r}
# Load in Gene Names and match up with rownames in ddsHTSeq
gene_names = read.table("~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/gene_names.txt",row.names=1,header = F)
colnames(gene_names)=c("GeneName","Source","BioType")
```
#Set-Up DGE Object
```{r}

dge=DGEList(counts=mergedcounts)

dge$samples$group=conds
dge$samples$batch=batch

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - Merged (Voom)', Rowv = F, RowSideColors = cond_colours[conds])

quartz()
par(mfrow=c(2,1))

barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data - Voom",ylab="Log-cpm", cex.axis = 0.5)

quartz()
plotMDS(dge, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)
```
  

#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds+batch)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)


#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds+batch)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

```{r}
#PCA Plot
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)]

percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")

voompca.df = data.frame(voompca.df, Condition = dge$samples$group, Batch = dge$samples$batch)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data-Voom)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-220, 220) +
  ylim(-220, 210)

#Tsne
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group, Batch=dge$samples$batch)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Merged Data-Voom)")+
  scale_color_manual(values=unique(cond_colours[conds]))+
  stat_ellipse(aes(colour=Type),type = "t")+
  xlim(-30,25)+
  ylim(-30,25)

```

```{r}
#Extract top significant hits by adjusted p value and log fold change
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Voom)')
  points(hit_table1[hitlist1[1:250],"logFC"],-log10(hit_table1[hitlist1[1:250],"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')
  

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', Colv=FALSE, col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = cond_colours[conds])

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = batch_colours[batch])

#Comparing distribution between TCGA and GTEX normals
quartz()
par(mfrow=c(2,2))
qqnorm(log2(dge$counts+1)[,1], main = 'Normal Q-Q Plot (GTEX Sample 1)')
qqnorm(log2(dge$counts+1)[,2], main = 'Normal Q-Q Plot (GTEX Sample 2)')
qqnorm(log2(dge$counts+1)[,595], main = 'Normal Q-Q Plot (TCGA Sample 1)')
qqnorm(log2(dge$counts+1)[,596], main = 'Normal Q-Q Plot (TCGA Sample 2)')

```

```{r}
hits=(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

up=hits[hits$logFC>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$logFC<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)

saveRDS(down, file = "~/Desktop/part2_project/Liver_Cancer/down25voom_down.RDS")
saveRDS(up, file = "~/Desktop/part2_project/Liver_Cancer/down25voom_up.RDS")

VOOMprimhits=as.data.frame(gene_names[hitlist1[1:250],]$GeneName)

```

```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/down_25.RData")

```











#33% Downsample

#Load in original gtex matrix and downsample by 33%
```{r}
gtex=readRDS('GTEX.rds')

gtex_samples=read.table('gtex_simple_samples.txt',row.names=1,sep="\t",header=F)
colnames(gtex_samples)=c('Tissue_Type', 'Description')

#Liver samples only - '-' substituted for '.' to match initial GTEX counts data
liver_ids = rownames(gtex_samples)[gtex_samples$Tissue_Type == "Liver"]
liver_ids = gsub("-",".",liver_ids)

gtex = gtex[,intersect(liver_ids,colnames(gtex))]

rownames(gtex)=gsub(".\\d+$","",rownames(gtex),perl=T)

gtex=as.matrix(gtex)
gtex33=downsampleMatrix(gtex, 0.67)
```

#Load in tcgarawcounts and merge the two
```{r}
tcgarawcounts=readRDS('~/Desktop/part2_project/Liver_Cancer/tcgarawcounts.RDS')

common_ids = intersect(rownames(gtex33),rownames(tcgarawcounts))

gtex33  = gtex33[common_ids, ]

rawcounts = tcgarawcounts[common_ids, ] 

p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250
```

#MERGE GTEX Normals & TCGA Primary Tumour Matrices
```{r}
mergedcounts=cbind(gtex33, rawcounts, deparse.level=1)
mergedcounts=mergedcounts[rowSums(mergedcounts)>=3,]

#Labelling columns by Sample Type
colnames(mergedcounts)[1:ncol(gtex33)]='Normal'

#Identify Conditions and Batches (GTEX vs. TCGA)
conds=as.factor(colnames(mergedcounts))
cond_colours = brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

batch = as.factor(c(rep("GTEX",ncol(gtex)),rep("TCGA",ncol(rawcounts))))
batch_colours=c('greenyellow', 'dark green')
names(batch_colours)=unique(batch)

hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
```

```{r}
# Load in Gene Names and match up with rownames in ddsHTSeq
gene_names = read.table("~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/gene_names.txt",row.names=1,header = F)
colnames(gene_names)=c("GeneName","Source","BioType")
```
#Set-Up DGE Object
```{r}

dge=DGEList(counts=mergedcounts)

dge$samples$group=conds
dge$samples$batch=batch

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - Merged (Voom)', Rowv = F, RowSideColors = cond_colours[conds])

quartz()
par(mfrow=c(2,1))

barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data - Voom",ylab="Log-cpm", cex.axis = 0.5)

quartz()
plotMDS(dge, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)
```
  

#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds+batch)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)


#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds+batch)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

```{r}
#PCA Plot
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)]

percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")

voompca.df = data.frame(voompca.df, Condition = dge$samples$group, Batch = dge$samples$batch)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data-Voom)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-220, 220) +
  ylim(-220, 210)

#Tsne
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group, Batch=dge$samples$batch)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Merged Data-Voom)")+
  scale_color_manual(values=unique(cond_colours[conds]))+
  stat_ellipse(aes(colour=Type),type = "t")+
  xlim(-30,30)+
  ylim(-25,30)

```

```{r}
#Extract top significant hits by adjusted p value and log fold change
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Voom)')
  points(hit_table1[hitlist1[1:250],"logFC"],-log10(hit_table1[hitlist1[1:250],"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')
  

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', Colv=FALSE, col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = cond_colours[conds])

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = batch_colours[batch])

#Comparing distribution between TCGA and GTEX normals
quartz()
par(mfrow=c(2,2))
qqnorm(log2(dge$counts+1)[,1], main = 'Normal Q-Q Plot (GTEX Sample 1)')
qqnorm(log2(dge$counts+1)[,2], main = 'Normal Q-Q Plot (GTEX Sample 2)')
qqnorm(log2(dge$counts+1)[,595], main = 'Normal Q-Q Plot (TCGA Sample 1)')
qqnorm(log2(dge$counts+1)[,596], main = 'Normal Q-Q Plot (TCGA Sample 2)')

```

```{r}
hits=(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

up=hits[hits$logFC>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$logFC<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)

saveRDS(down, file = "~/Desktop/part2_project/Liver_Cancer/down33voom_down.RDS")
saveRDS(up, file = "~/Desktop/part2_project/Liver_Cancer/down33voom_up.RDS")

VOOMprimhits=as.data.frame(gene_names[hitlist1[1:250],]$GeneName)

```


```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/down_33.RData")

```













#50% Downsample

#Load in original gtex matrix and downsample by 50%
```{r}
gtex=readRDS('GTEX.rds')

gtex_samples=read.table('gtex_simple_samples.txt',row.names=1,sep="\t",header=F)
colnames(gtex_samples)=c('Tissue_Type', 'Description')

#Liver samples only - '-' substituted for '.' to match initial GTEX counts data
liver_ids = rownames(gtex_samples)[gtex_samples$Tissue_Type == "Liver"]
liver_ids = gsub("-",".",liver_ids)

gtex = gtex[,intersect(liver_ids,colnames(gtex))]

rownames(gtex)=gsub(".\\d+$","",rownames(gtex),perl=T)

gtex=as.matrix(gtex)
gtex50=downsampleMatrix(gtex, 0.5)
```

#Load in tcgarawcounts and merge the two
```{r}
tcgarawcounts=readRDS('~/Desktop/part2_project/Liver_Cancer/tcgarawcounts.RDS')

common_ids = intersect(rownames(gtex50),rownames(tcgarawcounts))

gtex50  = gtex50[common_ids, ]

rawcounts = tcgarawcounts[common_ids, ] 

p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250
```

#MERGE GTEX Normals & TCGA Primary Tumour Matrices
```{r}
mergedcounts=cbind(gtex50, rawcounts, deparse.level=1)
mergedcounts=mergedcounts[rowSums(mergedcounts)>=3,]

#Labelling columns by Sample Type
colnames(mergedcounts)[1:ncol(gtex50)]='Normal'

#Identify Conditions and Batches (GTEX vs. TCGA)
conds=as.factor(colnames(mergedcounts))
cond_colours = brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

batch = as.factor(c(rep("GTEX",ncol(gtex)),rep("TCGA",ncol(rawcounts))))
batch_colours=c('greenyellow', 'dark green')
names(batch_colours)=unique(batch)

hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
```

```{r}
# Load in Gene Names and match up with rownames in ddsHTSeq
gene_names = read.table("~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/gene_names.txt",row.names=1,header = F)
colnames(gene_names)=c("GeneName","Source","BioType")
```

#Set-Up DGE Object
```{r}

dge=DGEList(counts=mergedcounts)

dge$samples$group=conds
dge$samples$batch=batch

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - Merged (Voom)', Rowv = F, RowSideColors = cond_colours[conds])

quartz()
par(mfrow=c(2,1))

barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data - Voom",ylab="Log-cpm", cex.axis = 0.5)

quartz()
plotMDS(dge, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)
```
  

#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds+batch)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)


#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds+batch)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

```{r}
#PCA Plot
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)]

percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")

voompca.df = data.frame(voompca.df, Condition = dge$samples$group, Batch = dge$samples$batch)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data-Voom)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-220, 220) +
  ylim(-220, 210)

#Tsne
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group, Batch=dge$samples$batch)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Merged Data-Voom)")+
  scale_color_manual(values=unique(cond_colours[conds]))+
  stat_ellipse(aes(colour=Type),type = "t")+
  xlim(-20,25)+
  ylim(-30,25)

```

```{r}
#Extract top significant hits by adjusted p value and log fold change
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Voom)')
  points(hit_table1[hitlist1[1:250],"logFC"],-log10(hit_table1[hitlist1[1:250],"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')
  

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', Colv=FALSE, col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = cond_colours[conds])

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = batch_colours[batch])

#Comparing distribution between TCGA and GTEX normals
quartz()
par(mfrow=c(2,2))
qqnorm(log2(dge$counts+1)[,1], main = 'Normal Q-Q Plot (GTEX Sample 1)')
qqnorm(log2(dge$counts+1)[,2], main = 'Normal Q-Q Plot (GTEX Sample 2)')
qqnorm(log2(dge$counts+1)[,595], main = 'Normal Q-Q Plot (TCGA Sample 1)')
qqnorm(log2(dge$counts+1)[,596], main = 'Normal Q-Q Plot (TCGA Sample 2)')

```

```{r}
hits=(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

up=hits[hits$logFC>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$logFC<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)

saveRDS(down, file = "~/Desktop/part2_project/Liver_Cancer/down50voom_down.RDS")
saveRDS(up, file = "~/Desktop/part2_project/Liver_Cancer/down50voom_up.RDS")

VOOMprimhits=as.data.frame(gene_names[hitlist1[1:250],]$GeneName)

```


```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/down_50.RData")

```












#50% Downsample

#Load in original gtex matrix and downsample by 50%
```{r}
gtex=readRDS('GTEX.rds')

gtex_samples=read.table('gtex_simple_samples.txt',row.names=1,sep="\t",header=F)
colnames(gtex_samples)=c('Tissue_Type', 'Description')

#Liver samples only - '-' substituted for '.' to match initial GTEX counts data
liver_ids = rownames(gtex_samples)[gtex_samples$Tissue_Type == "Liver"]
liver_ids = gsub("-",".",liver_ids)

gtex = gtex[,intersect(liver_ids,colnames(gtex))]

rownames(gtex)=gsub(".\\d+$","",rownames(gtex),perl=T)

gtex=as.matrix(gtex)
gtex75=downsampleMatrix(gtex, 0.25)
```

#Load in tcgarawcounts and merge the two
```{r}
tcgarawcounts=readRDS('~/Desktop/part2_project/Liver_Cancer/tcgarawcounts.RDS')

common_ids = intersect(rownames(gtex75),rownames(tcgarawcounts))

gtex75  = gtex75[common_ids, ]

rawcounts = tcgarawcounts[common_ids, ] 

p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250
```

#MERGE GTEX Normals & TCGA Primary Tumour Matrices
```{r}
mergedcounts=cbind(gtex75, rawcounts, deparse.level=1)
mergedcounts=mergedcounts[rowSums(mergedcounts)>=3,]

#Labelling columns by Sample Type
colnames(mergedcounts)[1:ncol(gtex75)]='Normal'

#Identify Conditions and Batches (GTEX vs. TCGA)
conds=as.factor(colnames(mergedcounts))
cond_colours = brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

batch = as.factor(c(rep("GTEX",ncol(gtex)),rep("TCGA",ncol(rawcounts))))
batch_colours=c('greenyellow', 'dark green')
names(batch_colours)=unique(batch)

hmcol = colorRampPalette(brewer.pal(9, 'GnBu'))(100)
```

```{r}
# Load in Gene Names and match up with rownames in ddsHTSeq
gene_names = read.table("~/Desktop/part2_project/Liver_Cancer/TCGA_Counts_mRNA_genelevel/gene_names.txt",row.names=1,header = F)
colnames(gene_names)=c("GeneName","Source","BioType")
```

#Set-Up DGE Object
```{r}

dge=DGEList(counts=mergedcounts)

dge$samples$group=conds
dge$samples$batch=batch

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - Merged (Voom)', Rowv = F, RowSideColors = cond_colours[conds])

quartz()
par(mfrow=c(2,1))

barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data (Downsampled 75%)', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data (Downsampled 75%)",ylab="Log-cpm", cex.axis = 0.5)

quartz()
plotMDS(dge, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)
```
  

#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds+batch)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)


#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds+batch)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

```{r}
#PCA Plot
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)]

percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")

voompca.df = data.frame(voompca.df, Condition = dge$samples$group, Batch = dge$samples$batch)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Downsampled 75%)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-220, 220) +
  ylim(-220, 210)

#Tsne
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group, Batch=dge$samples$batch)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Downsampled 75%)")+
  scale_color_manual(values=unique(cond_colours[conds]))+
  stat_ellipse(aes(colour=Type),type = "t")+
  xlim(-25,25)+
  ylim(-20,30)

```

```{r}
#Extract top significant hits by adjusted p value and log fold change
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Downsampled 75%)')
  points(hit_table1[hitlist1[1:250],"logFC"],-log10(hit_table1[hitlist1[1:250],"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')
  

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour (Downsampled 75%)', ColSideColors = cond_colours[conds])

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour (Downsampled 75%)', ColSideColors = batch_colours[batch])

#Comparing distribution between TCGA and GTEX normals
quartz()
par(mfrow=c(2,2))
qqnorm(log2(dge$counts+1)[,1], main = 'Normal Q-Q Plot (GTEX Sample 1)')
qqnorm(log2(dge$counts+1)[,2], main = 'Normal Q-Q Plot (GTEX Sample 2)')
qqnorm(log2(dge$counts+1)[,595], main = 'Normal Q-Q Plot (TCGA Sample 1)')
qqnorm(log2(dge$counts+1)[,596], main = 'Normal Q-Q Plot (TCGA Sample 2)')

```

```{r}
hits=(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

up=hits[hits$logFC>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$logFC<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)

saveRDS(down, file = "~/Desktop/part2_project/Liver_Cancer/down75voom_down.RDS")
saveRDS(up, file = "~/Desktop/part2_project/Liver_Cancer/down75voom_up.RDS")

VOOMprimhits=as.data.frame(gene_names[hitlist1[1:250],]$GeneName)

```


```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/down_75.RData")
```














##Venn Diagrams with downsampled hits

```{r}
tcga_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/tcgavoom_up.RDS')
tcga_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/tcgavoom_down.RDS')

merge_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/mergedvoom_up.RDS')
merge_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/mergedvoom_down.RDS')

gtex_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/gtexvoom_up.RDS')
gtex_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/gtexvoom_down.RDS')

down25_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down25voom_up.RDS')
down25_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down25voom_down.RDS')

down33_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down33voom_up.RDS')
down33_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down33voom_down.RDS')

down50_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down50voom_up.RDS')
down50_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down50voom_down.RDS')

down75_up=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down75voom_up.RDS')
down75_down=readRDS(file = '~/Desktop/part2_project/Liver_Cancer/updown/down75voom_down.RDS')

colnames(tcga_up)='GeneName'
colnames(tcga_down)='GeneName'
colnames(merge_up)='GeneName'
colnames(merge_down)='GeneName'
colnames(gtex_up)='GeneName'
colnames(gtex_down)='GeneName'

colnames(down25_up)='GeneName'
colnames(down25_down)='GeneName'
colnames(down33_up)='GeneName'
colnames(down33_down)='GeneName'
colnames(down50_up)='GeneName'
colnames(down50_down)='GeneName'
colnames(down75_up)='GeneName'
colnames(down75_down)='GeneName'


dev.off()

save.image(file = '~/Desktop/part2_project/Liver_Cancer/updown.Rdata')

```


```{r}

gtex_hittable=readRDS(file ='~/Desktop/part2_project/Liver_Cancer/GTEXhittable.RDS' )
tcga_hittable=readRDS(file='~/Desktop/part2_project/Liver_Cancer/TCGAhittable.RDS')

common_ids = intersect(rownames(gtex_hittable),rownames(tcga_hittable))

gtex_hittable=gtex_hittable[common_ids,]
tcga_hittable=tcga_hittable[common_ids,]

pdf(file='~/Desktop/part2_project/Liver_Cancer/plots/butterfly_plot.pdf')

plot(tcga_hittable$logFC,gtex_hittable$logFC, title(main = 'Butterfly Plot', xlab = 'Log2 Fold Change: TCGA Normals vs. Primary Tumour', ylab = 'Log2 Fold Change: GTEx Normals vs. Primary Tumour'), xlim=c(-10,10))
  abline(h=0)
  abline(v=0)
  abline(a=0,b=1, lty=1, col='red')

  dev.off()
```


