---
title: "Merged.Rmd"
author: "Tara Morrison"
date: "07/03/2019"
output: word_document
---


```{r}
setwd("~/Desktop/part2_project/Liver_Cancer")

library(RColorBrewer)
library(gplots)
library(ggplot2)
library(DESeq2)
library(Rtsne)
library(pca3d)
library(limma)
library(genefilter)
library(sva)
library(edgeR)
library(statmod)
library(reticulate)
library(VennDiagram)
```


## Load in GTEX Data
```{r}

gtex=readRDS('GTEX.rds')

# We obtained a simple table from the raw GTEX sample attributes using:
# cut -f 1,6,7 gtex_sample_attributes.txt  | sort -k2 > gtex_simple_samples.txt
# included: Sample IDs, Tissue, Tissue Description

#All GTEX Samples for all Tissue types
gtex_samples=read.table('gtex_simple_samples.txt',row.names=1,sep="\t",header=F)
colnames(gtex_samples)=c('Tissue_Type', 'Description')

#Liver samples only - '-' substituted for '.' to match initial GTEX counts data
liver_ids = rownames(gtex_samples)[gtex_samples$Tissue_Type == "Liver"]
liver_ids = gsub("-",".",liver_ids)

gtex = gtex[,intersect(liver_ids,colnames(gtex))]

#Need to remove decimal in gtex gene names after initial input

rownames(gtex)=gsub(".\\d+$","",rownames(gtex),perl=T)

common_ids = intersect(rownames(gtex),rownames(tcgarawcounts))

gtex  = gtex[common_ids, ]

#Subset GTEX results (20 samples)
#gtex = gtex[,1:20]
rawcounts = tcgarawcounts[common_ids, ] 

```

#Merging GTEX Data with TCGA Data
```{r}
#MERGE GTEX & TCGA MATRICES
mergedcounts=cbind(gtex, rawcounts, deparse.level=1)

#Labelling columns by Sample Type
colnames(mergedcounts)[1:ncol(gtex)]='Normal'
colnames(mergedcounts)[ncol(gtex)+1:ncol(rawcounts)]=paste(pdata$Sample_Type)

#Identify Conditions and Batches (GTEX vs. TCGA)
conds=as.factor(colnames(mergedcounts))
cond_colours = brewer.pal(length(unique(conds)), 'Set1')
names(cond_colours)=unique(conds)

batch = as.factor(c(rep("GTEX",ncol(gtex)),rep("TCGA",ncol(rawcounts))))
batch_colours=c('greenyellow', 'dark green')
names(batch_colours)=unique(batch)

```

#Save initial GTEX workspace
```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/merged_initial.RData")
```

##DESeq2

#ColData object produced - Inlcuding Batch Effect
```{r}

#Create colData object for DESeq
coldata=as.data.frame(conds)
coldata$batch=batch

colnames(coldata)=c("Sample_Type", 'Batch')

```



##DESeq Analysis

#DESeq Object from merged counts matrix 'mergedcounts'
```{r}
mergeddsHTSeq<-DESeqDataSetFromMatrix(countData=mergedcounts,colData=coldata,design=~Sample_Type+Batch)

mergedds=estimateSizeFactors(mergeddsHTSeq)
mergedds=estimateDispersions(mergedds)

```


#Data normalisation
```{r}
mergenormcounts <- counts(mergedds, normalized=TRUE)
mergerawcounts=counts(mergedds,normalized=FALSE)
mergelog2counts=log2(mergenormcounts+1)

quartz()
par(mfrow=c(2,1))

barplot(colSums(counts(mergedds, normalized=FALSE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Pre Normalised Counts')
legend("topright",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])

barplot(colSums(counts(mergedds, normalized=TRUE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Post Normalised Counts - DESeq2')
legend("topright",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])
```

#Disperison Pot
```{r}
mergedds=nbinomWaldTest(mergedds)
counts_table=counts(mergedds, normalized=TRUE)

quartz()
plotDispEsts(mergedds,main='Dispersion Plot - DESeq2')
```


#VST - merged data
```{r}
#blind added - to ignore variation due to batch effect
#If many of genes have large differences in counts due to the experimental design, it is important to set blind=FALSE for downstream analysis.

mergevsd <- varianceStabilizingTransformation(mergedds, blind = F)
mergevstMat <- assay(mergevsd)
mergevstcounts <- mergevstMat[order(apply(mergevstMat,1,sum),decreasing =TRUE),]

quartz()
heatmap.2(cor(mergevstcounts),trace="none",labRow = coldata$Sample_Type, cexRow = 0.6, labCol = coldata$Sample_Type, cexCol = 0.6, col=hmcol, main = 'Heatmap of VST - DESeq2', ColSideColors = cond_colours[conds])

```

#PCA Plots
```{r}
#Prcomp PCA
mergepca = prcomp(t(assay(mergevsd)))

mergepca.df = mergepca$x[, c(1,2,3)]

# Percentage explained by each component
percentages = round(mergepca$sdev^2 / sum(mergepca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(mergepca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")

mergepca.df = data.frame(mergepca.df, Condition = mergedds$Sample_Type, Batch = mergedds$Batch)

quartz()
ggplot(mergepca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data - DESeq)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-200,220) +
  ylim(-200, 210)

```

#Rtsne
```{r}
set.seed(72)
tsne = Rtsne(t(mergevstMat), perplexity = 30)
tsne.df=data.frame(tsne.1=tsne$Y[,1], tsne.2=tsne$Y[,2], Type=as.factor(mergedds$Sample_Type), Batch=as.factor(mergedds$Batch))

quartz()
ggplot(data = tsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  stat_ellipse(aes(colour=Type),type = "t")+
  ggtitle("R-tsne (Merged Data - DESeq)")+
  scale_color_manual(values=unique(cond_colours[tsne.df$Type]))+
  xlim(-30, 30) +
  ylim(-32, 30)
```


#Statistical Analysis
```{r}
p_threshold=0.01
lfc_threshold=1.5
max_display_hits=250

condition_list=levels(conds)

for (i in 2:(length(condition_list))){
 
  normal=condition_list[1]
  compare=condition_list[i]

  print(paste('(GTEX/TCGA Merged) Comparing', normal, 'vs.', compare))
  
  res=results(mergedds, contrast = c('Sample_Type', normal, compare))
  res=res[order(abs(res$log2FoldChange), decreasing = T),]
  
  hits=(rownames(res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
  
   print(paste("Found: ",length(hits)," Statistical hits"))
   
  if (length(hits) != 0){
    if (length(hits)>max_display_hits){
      hits=hits[1:max_display_hits]
    }
    
  quartz()
  plot(res$log2FoldChange,-log10(res$padj), ylab='-log10(Adjusted P-value)', xlab='log2 FoldChange', main=paste('Volcano Plot:', normal, 'vs.', compare,'(GTEX/TCGA Merged)'), pch=19, cex=0.2)
  points(res[hits,'log2FoldChange'],-log(res[hits,'padj'],10),pch=21,cex=0.4,col='turquoise1')
  text(res[hits[1:100],]$log2FoldChange,-log10(res[hits[1:100],]$padj), labels = gene_names[hits[1:100],]$GeneName, pos=3, cex = 0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-1.5, lty=3, col='red')
  abline(v=1.5, lty=3, col='red')
 
  quartz()
  heatmap.2(mergevstMat[hits,],trace='none',col=hmcol,dendrogram='col',main=paste("Heatmap Sig. Hits for", compare),labCol = conds, cexCol = 0.5, ColSideColors = cond_colours[conds])
  heatmap.2(mergevstMat[hits,],trace='none',col=hmcol,dendrogram='col',main=paste("Heatmap Sig. Hits for", compare),labCol = conds, cexCol = 0.5, ColSideColors = batch_colours[batch])
  } else {
    print(paste("No Results for ",compare))
  }
}
```

#Medians
```{r}
median.primary=apply(mergevstMat[,176:546],1,median)
median.normal=apply(mergevstMat[,c(1:175, 547:596)], 1, median)

quartz()
plot(median.normal,median.primary, main="Normal vs. Primary_Tumor (GTEX/TCGA Merged)", pch=19, col='darkblue', cex=0.4)
points(median.normal[hits],median.primary[hits], pch=21, col='turquoise1')
text(median.normal[hits],median.primary[hits],cex=0.4,pos=4,labels = gene_names[hits,]$GeneName)
abline(a=0,b=1, col='red', lwd=2, cex=0.4, lty=2)
```

#Up- and down-regulated significant gene lists
```{r}
hits=((res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
up=hits[hits$log2FoldChange>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$log2FoldChange<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)
```

#Save out workspace
```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/merged_DESeq.RData")
```


##Voom

#Set-Up DGE Object
```{r}
dge=DGEList(counts=mergedcounts)

dge$samples$group=conds
dge$samples$batch=batch

cpm=cpm(dge)
lcpm=cpm(dge, log=T)

keep.exprs=rowSums(cpm>1)>=3
dge=dge[keep.exprs,,keep.lib.sizes=F]
dim(dge)

dge=calcNormFactors(dge, method='TMM')
dge$samples$norm.factors

quartz()

heatmap.2(cor(dge$counts),trace="none",labRow = conds, cexRow = 0.6, labCol =  dge$samples$group, cexCol = 0.6, col=hmcol, main = 'Heatmap of Correlation of the dge Counts - Merged (Voom)', Rowv = F, RowSideColors = cond_colours[conds])

quartz()
par(mfrow=c(2,1))

barplot(apply(dge$counts,2,sum), las=2, main='Pre Normalised Data', ylab='Log-cpm', cex.axis = 0.5, col = cond_colours[conds])

barplot(apply(dge$counts,2,sum)*dge$sample$norm.factors, las=2, col=cond_colours[conds], main="Normalised Data - Voom",ylab="Log-cpm", cex.axis = 0.5)

quartz()
plotMDS(dge, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)
```
  
#Design including conditions and batch effect
```{r}
#Design
design=model.matrix(~0+conds+batch)
contr.matrix=makeContrasts(condsPrimary_Tumor-condsNormal, levels = colnames(design))

quartz()
v=voom(dge, design, plot = T)

vfit=lmFit(v, design)
vfit=contrasts.fit(vfit, contrasts = contr.matrix)
efit=eBayes(vfit)


#SA Plot
quartz()
plotSA(efit, main='Mean-Variance Trend - Voom')


#MDS Plot
quartz()
plotMDS(v, pch = 19, col= cond_colours[conds], main='Multi Dimensional Scaling (MDS) Plot - Voom (conds+batch)')
legend('bottomright', legend=levels(conds), col=cond_colours[unique(conds)], pch = 20)


#Summary of findings from comparisons
summary(decideTests(efit))

#Further Analysis
tfit = treat(efit, lfc=lfc_threshold)
dt = decideTests(tfit)
summary(dt)
```

#PCA Plot
```{r}
voompca = prcomp(t(v$E))

voompca.df = voompca$x[, c(1,2,3)]
# Percentage explained by each component
percentages = round(voompca$sdev^2 / sum(voompca$sdev^2) * 100)[1:3]
percentages = paste(paste0(colnames(voompca.df), ":"),  paste0(as.character(percentages), "%"), "variance", sep=" ")
voompca.df = data.frame(voompca.df, Condition = dge$samples$group, Batch = dge$samples$batch)

quartz()
ggplot(voompca.df, aes(x=PC1,y=PC2, colour = Condition, shape = Batch))+
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data-Voom)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-220, 220) +
  ylim(-220, 210)
```

#Tsne
```{r}
voomtsne = Rtsne(t(cpm), perplexity = 30)
voomtsne.df=data.frame(tsne.1=voomtsne$Y[,1], tsne.2=voomtsne$Y[,2], Type=dge$samples$group, Batch=dge$samples$batch)

quartz()
ggplot(data = voomtsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Merged Data-Voom)")+
  scale_color_manual(values=unique(cond_colours[conds]))+
  stat_ellipse(aes(colour=Type),type = "t")+
  xlim(-25,25)+
  ylim(-30,25)
```

#Extract top significant hits by adjusted p value and log fold change
```{r}
hit_table1=topTable(efit,coef=1,number=10000000000000)

hit_table1=hit_table1[order(abs(hit_table1$logFC),decreasing = T),]

hitlist1=rownames(hit_table1[((abs(hit_table1$logFC) >= lfc_threshold) & (hit_table1$adj.P.Val<=p_threshold)),])

quartz()
plot(hit_table1$logFC,-log10(hit_table1$adj.P.Val),pch=19,cex=0.4,col="darkblue", main='Volcano Plot - Normal vs Primary Tumour (Voom)')
  points(hit_table1[hitlist1[1:250],"logFC"],-log10(hit_table1[hitlist1[1:250],"adj.P.Val"]),cex=0.4,col="cyan")
  text(hit_table1[1:100,]$logFC,-log10(hit_table1[1:100,]$adj.P.Val), labels= gene_names[hitlist1[1:100],]$GeneName, pos=3, cex=0.4)
  abline(h=-log10(p_threshold), lty=3, col='red')
  abline(v=-lfc_threshold, lty=3, col='red')
  abline(v=lfc_threshold, lty=3, col='red')
  

# Heatmap the top 100 from above using the log 2 of the voom normalised counts.
quartz()

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = cond_colours[conds])

heatmap.2(lcpm[hitlist1[1:100],], trace = 'none', col=hmcol, main = 'Heatmap of lcpm for Primary Tumour', ColSideColors = batch_colours[batch])
```

```{r}
#Comparing distribution between TCGA and GTEX normals
quartz()
par(mfrow=c(2,2))
qqnorm(log2(dge$counts+1)[,1], main = 'Normal Q-Q Plot (GTEX Sample 1)')
qqnorm(log2(dge$counts+1)[,2], main = 'Normal Q-Q Plot (GTEX Sample 2)')
qqnorm(log2(dge$counts+1)[,595], main = 'Normal Q-Q Plot (TCGA Sample 1)')
qqnorm(log2(dge$counts+1)[,596], main = 'Normal Q-Q Plot (TCGA Sample 2)')

apply(dge$counts, 2, median)

```

Save out workspace
```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/merged_edger.RData")
```


##SVA

#Design with surrogate variables
```{r}
#Removing hidden batch effects using SVA
DESeqdds=DESeq(mergedds)
dat=counts(DESeqdds,normalized=T)
idx=rowMeans(dat)>1
dat=dat[idx,]

mod=model.matrix(~Sample_Type, colData(DESeqdds))
mod0=model.matrix(~1, colData(DESeqdds))

n.sv = num.sv(dat,mod,method="leek") 
save.image("~/Desktop/part2_project/Liver_Cancer/merged_svaNSV.RData")

svseq=svaseq(dat, mod, mod0, n.sv=592)

quartz()
par(mfrow=c(2,1),mar=c(3,5,3,1))

stripchart(svseq$sv[,1] ~ DESeqdds$Batch,vertical=TRUE,main="SV1")
abline(h=0)
stripchart(svseq$sv[,1] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV1")
abline(h=0)

stripchart(svseq$sv[,2] ~ DESeqdds$Batch,vertical=TRUE,main="SV2")
abline(h=0)
stripchart(svseq$sv[,2] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV2")
abline(h=0)

stripchart(svseq$sv[,3] ~ DESeqdds$Batch,vertical=TRUE,main="SV3")
abline(h=0)
stripchart(svseq$sv[,3] ~ DESeqdds$Sample_Type,vertical=TRUE,main="SV3")
abline(h=0)


ddssva = DESeqdds

ddssva$SV1= svseq$sv[,1] 

design(ddssva)= ~ SV1 + Sample_Type
```

#Data normalisation
```{r}
svanormcounts <- counts(ddssva, normalized=TRUE)
svarawcounts=counts(ddssva,normalized=FALSE)
svalog2counts=log2(svanormcounts+1)

quartz()
par(mfrow=c(2,2))

barplot(colSums(counts(ddssva, normalized=FALSE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Pre Normalised Counts')
legend("topright",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])

barplot(colSums(counts(ddssva, normalized=TRUE)), col=cond_colours[conds], las=2,cex.names=0.5,main='Post Normalised Counts - SVA')
legend("topleft",levels(conds),cex=0.5,fill=cond_colours[levels(conds)])

```

```{r}
ddssva=nbinomWaldTest(ddssva)

counts_table=counts(dds, normalized=TRUE)

quartz()
plotDispEsts(dds,main='Dispersion Plot - TCGA')
```

#VST
```{r}
svavsd <- varianceStabilizingTransformation(ddssva, blind = F)
svavstMat <- assay(svavsd)
svavstcounts <- svavstMat[order(apply(svavstMat,1,sum),decreasing =TRUE),]

quartz()
heatmap.2(cor(svavstcounts),trace="none",labRow = coldata$Sample_Type, cexRow = 0.6, labCol = coldata$Sample_Type, cexCol = 0.6, col=hmcol, main = 'Heatmap of VST - SVA', ColSideColors = cond_colours[conds])
```

#PCA Plots
```{r}
svapca = prcomp(t(assay(svavsd)))

svapca.df = svapca$x[, c(1,2,3)]

# Percentage explained by each component
percentages = round(svapca$sdev^2 / sum(pca$sdev^2) * 100)[1:3]
percentages = paste(paste(colnames(svapca.df), ":"),  paste(as.character(percentages), "%"), "variance", sep=" ")

svapca.df = data.frame(svapca.df, Condition = ddssva$Sample_Type, Batch = ddssva$Batch, SV = ddssva$SV1)

quartz()
ggplot(svapca.df, aes(x=PC1,y=PC2, colour = Condition ,  shape = Batch )) + 
	geom_point(size = 4)+
  scale_color_manual(values=unique(cond_colours[conds]))+
  ggtitle("PCA1 vs PCA2 (Merged Data-SVA)")+
  stat_ellipse(aes(colour=Condition),type = "t")+
  xlab(percentages[1]) + 
	ylab(percentages[2]) +
  xlim(-200, 220) +
  ylim(-200, 210)
```

#Tsne
```{r}
svatsne = Rtsne(t(svavstMat), perplexity = 30)
svatsne.df=data.frame(tsne.1=svatsne$Y[,1], tsne.2=svatsne$Y[,2], Type=as.factor(ddssva$Sample_Type), Batch=as.factor(ddssva$Batch))

quartz()
ggplot(data = svatsne.df, aes(tsne.1,  tsne.2, colour=Type, shape=Batch))+
  geom_point(size=4)+
  ggtitle("T-sne (Merged Data-SVA)")+
  stat_ellipse(aes(colour=Type),type = "t")+
  scale_color_manual(values=unique(cond_colours[tsne.df$Type])) +
  xlim(-30, 30) +
  ylim(-25, 25)
```

#Statistical Analysis - SVA
```{r}
condition_list=levels(conds)

for (i in 2:(length(condition_list))){
 
  normal=condition_list[1]
  compare=condition_list[2]

  print(paste('(SVA) Comparing', normal, 'vs.', compare))
  
  res=results(ddssva, contrast = c('Sample_Type', normal, compare))
  res=res[order(abs(res$log2FoldChange), decreasing = T),]
  
  hits=(rownames(res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
  
   print(paste("Found: ",length(hits)," Statistical hits"))
   
  if (length(hits) != 0){
    if (length(hits)>max_display_hits){
      hits=hits[1:max_display_hits]
    }
    
  quartz()
  plot(res$log2FoldChange,-log10(res$padj), ylab='-log10(Adjusted P-value)', xlab='log2 FoldChange', main=paste('Volcano Plot:', normal, 'vs.', compare,'(SVA)'), pch=19, cex=0.2)
  points(res[hits,'log2FoldChange'],-log(res[hits,'padj'],10),pch=21,cex=0.4,col='turquoise1')
  text(res[hits[1:100],]$log2FoldChange,-log10(res[hits[1:100],]$padj), labels = gene_names[hits[1:100],]$GeneName, pos=3, cex = 0.4)
  abline(h=-log10(0.05), lty=3, col='red')
  abline(v=-1, lty=3, col='red')
  abline(v=1, lty=3, col='red')

  
quartz()
  heatmap.2(svavstMat[hits,],trace='none',col=hmcol, dendrogram='col',main=paste("Heatmap Sig. Hits for", compare),labCol = conds, cexCol = 0.5, ColSideColors = cond_colours[conds]) 
  heatmap.2(svavstMat[hits,],trace='none',col=hmcol,dendrogram='col',main=paste("Heatmap Sig. Hits for", compare),labCol = conds, cexCol = 0.5, ColSideColors = batch_colours[batch])
  
  } else {
    print(paste("No Results for ",compare))
  }
  
}
```

#Up- and Down-regulated significant gene lists
```{r}
hits=((res[((res$padj<=p_threshold) & (abs(res$log2FoldChange)>=lfc_threshold) & (!is.na(res$padj))),]))
up=hits[hits$log2FoldChange>0,]
up=rownames(up)
up=as.data.frame(gene_names[up,]$GeneName)
down=hits[hits$log2FoldChange<0,]
down=rownames(down)
down=as.data.frame(gene_names[down,]$GeneName)

saveRDS(down, file = "~/Desktop/part2_project/Liver_Cancer/mergesva_down.RDS")
saveRDS(up, file = "~/Desktop/part2_project/Liver_Cancer/mergesva_up.RDS")
```

#save workspace
```{r}
save.image("~/Desktop/part2_project/Liver_Cancer/merged_sva.RData")
```

